---
title: "Project1"
format: html
editor: visual
---

# Getting data

To start, I'm going to put my API key here for safe keeping:

```{r}
API_key<-"640ce0de32b08c9ded3cc11874a33ea9200c2c78"
```




# Part 1: Data Processing

## Parsing Variable Values


The first thing I'm going to do is pull the list of variable names from the website, and parse them individually, starting with 2021.

```{r}
#| message: false
#| warning: false
#| echo: false
library(httr)
library(jsonlite)
library(tidyverse)
cen_vars21<-GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars21<-fromJSON(rawToChar(cen_vars21$content))
```


```{r}
vars21<-parsed_vars21$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

Now to do 2022 I can just change a couple of 1's to 2's: 

```{r}
cen_vars22<-GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars22<-fromJSON(rawToChar(cen_vars22$content))
```


```{r}
vars22<-parsed_vars22$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

For 2023 and 2024, though, I need to change the numbers, but also "ST" to "STATE":

```{r}
cen_vars23<-GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars23<-fromJSON(rawToChar(cen_vars23$content))
```


```{r}
vars23<-parsed_vars23$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

```{r}
cen_vars24<-GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars24<-fromJSON(rawToChar(cen_vars24$content))
```

```{r}
vars24<-parsed_vars24$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

Alright, so I have my lists, now to save them in an .rds file: 

```{r}
var_lists<-list(vars21,vars22,vars23,vars24)
saveRDS(var_lists,file="data/var_lists")
```

