---
title: "Project 1"
format: html
editor: visual
---

# Getting data

To start, I'm going to put my API key here for safe keeping:

```{r}
API_key<-"640ce0de32b08c9ded3cc11874a33ea9200c2c78"
```




# Part 1: Data Processing

## Parsing Variable Values


The first thing I'm going to do is pull the list of variable names from the website, and parse them individually, starting with 2021.

```{r}
#| message: false
#| warning: false
#| echo: false
library(httr)
library(jsonlite)
library(tidyverse)
cen_vars21<-GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars21<-fromJSON(rawToChar(cen_vars21$content))
```


```{r}
vars21<-parsed_vars21$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

Now to do 2022 I can just change a couple of 1's to 2's: 

```{r}
cen_vars22<-GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars22<-fromJSON(rawToChar(cen_vars22$content))
```


```{r}
vars22<-parsed_vars22$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

For 2023 and 2024, though, I need to change the numbers, but also "ST" to "STATE":

```{r}
cen_vars23<-GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars23<-fromJSON(rawToChar(cen_vars23$content))
```


```{r}
vars23<-parsed_vars23$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

```{r}
cen_vars24<-GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars24<-fromJSON(rawToChar(cen_vars24$content))
```

```{r}
vars24<-parsed_vars24$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

Alright, so I have my lists, now to save them in an .rds file: 

```{r}
var_lists<-list(vars21,vars22,vars23,vars24)
saveRDS(var_lists,file="data/var_lists")
```

## Parsing API Data

### First bullet

First up, I'll try to query the API using the normal method. Note that I kept the filtering from the given example to keep things from taking forever. 

```{r}
test<-GET("https://api.census.gov/data/2021/acs/acs1/pums?get=AGEP,SEX,ST&for=state:*&SCHL=24&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
test
```

```{r}
test_parsed<-fromJSON(rawToChar(test$content))
test_data<-as_tibble(test_parsed)
```

```{r}
head(test_data)
```


So, it looks like what we got back has the names for the variables in the first row, so we should fix that. 

```{r}
test_data_fixed<-test_data|>
  rename("AGEP"="V1","SEX"="V2","ST"="V3","SCHL"="V4","STATE"="V5")|>
  slice(-1)
test_data_fixed
```

Looks pretty good, I think, but what I did won't translate nicely into the function I need to write next. 

### Second bullet

Now, we need to work on getting a function that takes what's returned by GET() abd turns it into a nice tibble Note that I also found the "set_names" function online, which means I can fix the column names all at once (instead of piecemeal).

```{r}
make_tib<-function(x){
  query_parsed<-fromJSON(rawToChar(x$content))
  query_data<-as_tibble(query_parsed)|>
    set_names(query_parsed[1,])|>
    slice(-1)
query_data  
}

```

Note that test is still a defined GET() output from before. Let's take it for a spin.

```{r}
make_tib(test)
```

Seems alright!

### Third bullet

Now, the big one. We need to write a function that takes, as inputs, year of survey (default of 2024), numeric variables (default to AGEP, with PWGTP always returned), categorical variables (at least one, with SEX as the default), geography level (default to STATE) and one specific value of the variable.

I'm going to start by naming the function and setting up the function defaults and creating error messages if/when improper calls are made. For now, I'm going to just make the function add 1 to a number so I can focus on the function setup (with a focus on functionality to come).

```{r}
make_query<-function(x,year=2024,num_vars="AGEP",cat_vars="SEX",geo_level="STATE",geo_vars="everything"){
  if(year!=2024 & year!=2023 & year != 2022 & year != 2021){
    stop("You must enter year value of 2021, 2022, 2023 or use default of 2024.")
  }
  if(sum(num_vars %in% c("AGEP", "GASP", "GRPIP", "JWAP","JWDP"))!=length(num_vars) ){
    stop("You must enter a subset of the following numerical variables: AGEP, GASP, GRPIP, JWAP, JWDP.")
  }
  if(sum(cat_vars %in% c("FER", "HHL", "SCH", "SCHL","SEX"))!=length(cat_vars) ){
    stop("You must enter a subset of the following categorical variables: FER, HHL, SCH, SCHL,SEX.")
  }
  if((year == 2021 | year == 2022) & (geo_level %in% c("ST","REGION","DIVISION") != 1)){
    stop("You must enter select of the following geographical levels: ST (for year 2021 or 2022)/ STATE (for tear 2023 or 2024), REGION or DIVISION.")
  }
  if((year == 2023 | year == 2024) & (geo_level %in% c("STATE","REGION","DIVISION") != 1)){
    stop("You must enter select of the following geographical levels: ST (for year 2021 or 2022)/ STATE (for tear 2023 or 2024), REGION or DIVISION.")
  }
  x+1
}
```

I've done a fair amount of testing with the above conditional statements and STOP commands and I think I'm ready to have the function actually do what it's meant to do. 