---
title: "Project 1"
format: html
editor: visual
---

# Getting data

To start, I'm going to put my API key here for safe keeping:

```{r}
API_key<-"640ce0de32b08c9ded3cc11874a33ea9200c2c78"
```




# Part 1: Data Processing

## Parsing Variable Values


The first thing I'm going to do is pull the list of variable names from the website, and parse them individually, starting with 2021.

```{r}
#| message: false
#| warning: false
#| echo: false
library(httr)
library(jsonlite)
library(tidyverse)
cen_vars21<-GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars21<-fromJSON(rawToChar(cen_vars21$content))
```


```{r}
vars21<-parsed_vars21$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

Now to do 2022 I can just change a couple of 1's to 2's: 

```{r}
cen_vars22<-GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars22<-fromJSON(rawToChar(cen_vars22$content))
```


```{r}
vars22<-parsed_vars22$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

For 2023 and 2024, though, I need to change the numbers, but also "ST" to "STATE":

```{r}
cen_vars23<-GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars23<-fromJSON(rawToChar(cen_vars23$content))
```


```{r}
vars23<-parsed_vars23$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

```{r}
cen_vars24<-GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars24<-fromJSON(rawToChar(cen_vars24$content))
```

```{r}
vars24<-parsed_vars24$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

t this point, for convenience later, I'm going to rename the state variables in vars21 and vars22 to match the other years.

```{r}
vars21$STATE<-vars21$ST
vars22$STATE<-vars22$ST
```

Now, to avoid confusion, I'll remove the "ST" lists from vars21 & vars22:

```{r}
vars21$ST<-NULL
vars22$ST<-NULL
```

Alright, so I have my lists, now to save them in an .rds file: 

```{r}
var_lists<-list(vars21=vars21,vars22=vars22,vars23=vars23,vars24=vars24)
saveRDS(var_lists,file="data/var_lists")
```

## Parsing API Data

### First bullet

First up, I'll try to query the API using the normal method. Note that I kept the filtering from the given example to keep things from taking forever. 

```{r}
test<-GET("https://api.census.gov/data/2021/acs/acs1/pums?get=AGEP,SEX,ST&for=state:*&SCHL=24&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
test
```

```{r}
test_parsed<-fromJSON(rawToChar(test$content))
test_data<-as_tibble(test_parsed,.name_repair = "unique")
```
I got an error stating that the I must have unique column names and I found help related to name_repair online to fix that. Now, to look at what's coming out of this...

```{r}
head(test_data)
```


So, it looks like what we got back has the names for the variables in the first row, so we should fix that. 

```{r}
test_data_fixed<-test_data|>
  rename("AGEP"="...1","SEX"="...2","ST"="...3","SCHL"="...4","STATE"="...5")|>
  slice(-1)
test_data_fixed
```

Looks pretty good, I think, but what I did won't translate nicely into the function I need to write next. 

### Second bullet

Now, we need to work on getting a function that takes what's returned by GET() and turns it into a nice tibble Note that I also found the "set_names" function online, which means I can fix the column names all at once (instead of piecemeal).

```{r}
make_tib<-function(x){
  query_parsed<-fromJSON(rawToChar(x$content))
  query_data<-as_tibble(query_parsed)|>
    set_names(query_parsed[1,])|>
    slice(-1)
query_data  
}

```

Note that test is still a defined GET() output from before. Let's take it for a spin.

```{r}
make_tib(test)
```

Seems alright!

### Third bullet

Now, the big one. We need to write a function that takes, as inputs, year of survey (default of 2024), numeric variables (default to AGEP, with PWGTP always returned), categorical variables (at least one, with SEX as the default), geography level (default to STATE) and one specific value of the variable.

I'm going to start by  setting up the function defaults and creating error messages if/when improper calls are made. For now, I'm going to just make the function add 1 to a number so I can focus on the function setup (with a focus on functionality to come).

```{r}
query_validation<-function(yr=2024,num_vars="AGEP",cat_vars="SEX",geo_level="STATE",geo_vars="All",x=5){
  if(yr!=2024 & yr!=2023 & yr != 2022 & yr != 2021){
    stop("year error: You must enter year value of 2021, 2022, 2023 or use default of 2024.")
  }
  if(sum(num_vars %in% c("AGEP", "GASP", "GRPIP", "JWAP","JWDP"))!=length(num_vars) ){
    stop("num_vars error: You must enter a subset of the following numerical variables: AGEP, GASP, GRPIP, JWAP, JWDP.")
  }
  if(sum(cat_vars %in% c("FER", "HHL", "SCH", "SCHL","SEX"))!=length(cat_vars) || length(cat_vars)==0){
    stop("car_vars error: You must enter a nonempty subset of the following categorical variables: FER, HHL, SCH, SCHL,SEX.")
  }
  if(length(geo_level)>1 || geo_level %in% c("STATE","REGION","DIVISION") != 1){
    stop("geo_level error: You must select one of the following geographical levels: STATE, REGION or DIVISION.")
  }
  if(geo_level == "REGION" & !"All" %in% geo_vars & !all(geo_vars %in% names(vars24$REGION$values$item))){
    print(unlist(vars24$REGION$values$item))
    stop("geo_var error: You must select one or more of the regions listed above. Be sure to use the provided (numeric) label.For example, set geo_vars=c(1,2) for Northeast and Midwest.")
  }
  if(geo_level == "STATE" & !"All" %in% geo_vars & !all(geo_vars %in% unlist(vars24$STATE$values$item))){
    print(unlist(vars24$STATE$values$item))
    stop("geo_var error: You must select one or more of the states from the list above. Be sure to use the provided (numeric) label.For example, set geo_vars=c(1,2) for Alabama and Alaska.")
  }
  if(geo_level == "DIVISION" & !"All" %in% geo_vars & !all(geo_vars %in% unlist(vars24$DIVISION$values$item))){
    print(unlist(vars24$DIVISION$values$item))
    stop("geo_var error: You must select one or more of the divisions from the list above. Be sure to use the provided (numeric) label. For example, set geo_vars=c(0,1) for Puerto Rico and New England (Northeast region).")
  }
  x+1
}
```

Alright, it looks like all my validation stuff is working (at least, the logical tests are), so I'll hopefully be able to call it in my main query function. I'll start by making my function slowly, by first putting it together with a carefully curated URL, then add in the validation and tibble-making functions. Note that I'm going to start with the same defaults as I have for the validation function so, hopefully all the variables but 'x' get passed in nicely. 

It's worth noting here that I had a really hard time getting this to work out. For example, I initially ran into trouble because, when I wanted multiple  numerical and/or categorical variables, it would create a new URL for each value in the variable. It also took me quite a while to get the ability to pick geography level, and then either all or a subset of the variable values within the level selected. I amended the error code above to provided the lists the user could use to find the labels they need. As you can see below, the paste() and paste() functions where handy here when it came time to tie all together in the url for GET().

```{r}
make_query<-function(yr=2024,num_vars="AGEP",cat_vars="SEX",geo_level="STATE",geo_vars="All"){
  if((yr==2021 || yr==2022) && ("STATE" %in% geo_level)){
    geo_level<-"state"
  }
  if("ALL" %in% geo_vars){
    geo<-paste0("&for=",geo_level,":*")
  }else{
    geo<-paste0("&for=",geo_level,":",paste(geo_vars,collapse=","))
  }
  link<-paste0("https://api.census.gov/data/",yr,"/acs/acs1/pums?get=",paste(num_vars,collapse=","),",",paste(cat_vars,collapse=","),",PWGTP",geo,"&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
  GET(link)
}
```

I think it's working. *Crosses fingers real tight.* 

Next up, I"m going to expand my function to, first, run the validation (which should stop the whole shebang if it fails), then construct the link, then tibble-fy it all using the functions from before.  

```{r}
make_query<-function(yr=2024,num_vars="AGEP",cat_vars="SEX",geo_level="STATE",geo_vars="All"){
  query_validation(yr,num_vars,cat_vars,geo_level,geo_vars,x=5)
  if((yr==2021 || yr==2022) && ("STATE" %in% geo_level)){
    geo_level<-"state"
  }
  if("All" %in% geo_vars){
    geo<-paste0("&for=",geo_level,":*")
  }else{
    geo<-paste0("&for=",geo_level,":",paste(geo_vars,collapse=","))
  }
  link<-paste0("https://api.census.gov/data/",yr,"/acs/acs1/pums?get=",paste(num_vars,collapse=","),",",paste(cat_vars,collapse=","),",PWGTP",geo,"&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
  GET(link)|>
    make_tib()
}
```
Now, to test it a bit:

```{r}
make_query(geo_level="REGION",num_vars = c("GRPIP","GASP","JWAP"),yr=2021,geo_vars = 4)
```
 Hurray! Now I need to start working on the columns. 
