---
title: "Project1"
format: html
editor: visual
---

# Getting data

To start, I'm going to put my API key here for safe keeping: 

```{r}
API_key<-"640ce0de32b08c9ded3cc11874a33ea9200c2c78"
```

Now, to get our data from the Public Use Microdata Sample Census API for 2021 and, once that's working, adjust for the other years. 

```{r}
#| message: false
#| warning: false
#| echo: false
library(httr)
library(jsonlite)
library(tidyverse)
cen_data21<-GET("https://api.census.gov/data/2021/acs/acs1/pums?get=AGEP,GASP,GRPIP,JWAP,JWDP,JWMNP,PWGTP,FER,HHL,SCH,SCHL,SEX,ST,REGION,DIVISION&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
```

Now that I have that, I'll try to start parsing it: 

```{r}
parsed21<-fromJSON(rawToChar(cen_data21$content))
cen21_info<-as_tibble(parsed21$data)
cen21_info
```

When running this, I see that the first row is actually the variable names so I need to pull those out. Once I do that I'll need to remove the top row.
"
```{r}
cen21_info<-cen21_info|>
rename("AGEP"="V1","GASP"="V2","GRPIP"="V3","JWAP"="V4","JWDP"="V5","JWMNP"="V6","PWGTP"="V7","FER"="V8","HHL"="V9","SCH"="V10","SCHL"="V11","SEX"="V12","ST"="V13","REGION"="V14","DIVISION"="V15")|>
  slice(-1)
cen21_info
```

Great! Now that I have the column names sorted, I can remove the top row of my tibble. 



```{r}
cen21_info
```

