---
title: "Project 1"
format: html
editor: visual
---

# Getting data

To start, I'm going to put my API key here for safe keeping:

```{r}
##Saving my key for easy retreival. 
## 640ce0de32b08c9ded3cc11874a33ea9200c2c78
```


# Part 1: Data Processing

## Parsing Variable Values


The first thing I'm going to do is pull the list of variable names from the website, and parse them individually, starting with 2021. Mostly, I did it this way to get things figured out on a small scale before moving to the rest of the years. 

I'll also use this block to load the libraries I'll be using and set the option I need but can't seem to fix globally.

```{r}
#| message: false
#| warning: false
#| echo: false
library(httr)
library(jsonlite)
library(tidyverse)
library(lubridate)
library(hms)
```

```{r}
options(error = NULL)
```

```{r}
#Get first set of variables.
cen_vars21<-GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars21<-fromJSON(rawToChar(cen_vars21$content))
```


```{r}
# Save the variables I want to keep.
vars2021<-parsed_vars21$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

Now to do 2022 I can just change a couple of 1's to 2's: 

```{r}
# Repeating the above for 2022.
cen_vars22<-GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars22<-fromJSON(rawToChar(cen_vars22$content))
```


```{r}
# Ditto
vars2022<-parsed_vars22$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","ST","REGION","DIVISION")]
```

For 2023 and 2024, though, I need to change the numbers, but also "ST" to "STATE":

```{r}
# Repeating the above for 2023.
cen_vars23<-GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars23<-fromJSON(rawToChar(cen_vars23$content))
vars2023<-parsed_vars23$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

```{r}
# Repeating the above for 2024.
cen_vars24<-GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json?key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
parsed_vars24<-fromJSON(rawToChar(cen_vars24$content))
vars2024<-parsed_vars24$variables[c("AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP","PWGTP","FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION")]
```

At this point, for convenience later, I'm going to rename the state variables in vars2021 and vars2022 to match the other years.

```{r}
#Renaming STATE variables
vars2021$STATE<-vars2021$ST
vars2022$STATE<-vars2022$ST

# To avoid confusion, I'll remove the "ST" lists from vars2021 & vars2022:

vars2021$ST<-NULL
vars2022$ST<-NULL
```

Alright, so I have my lists, now to save them in an .rds file: 

```{r}
# Saving lists to RDS
var_lists<-list(vars2021=vars2021,vars2022=vars2022,vars2023=vars2023,vars2024=vars2024)
saveRDS(var_lists,file="data/var_lists")
```

## Parsing API Data

### First bullet

First up, I'll try to query the API using the normal method. Note that I kept the filtering from the given example to keep things from taking forever. 

```{r}
#Trial API pull
test<-GET("https://api.census.gov/data/2021/acs/acs1/pums?get=AGEP,SEX,ST&for=state:*&SCHL=24&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
test
```

```{r}
# Parsing result, fixing naming issue
test_parsed<-fromJSON(rawToChar(test$content))
test_data<-as_tibble(test_parsed,.name_repair = "unique")
```
I got an error stating that the I must have unique column names and I found help related to name_repair online to fix that. Now, to look at what's coming out of this...

```{r}
#Examine result
head(test_data)
```


So, it looks like what we got back has the names for the variables in the first row, so we should fix that. 

```{r}
#Renaming variable names
test_data_fixed<-test_data|>
  rename("AGEP"="...1","SEX"="...2","ST"="...3","SCHL"="...4","STATE"="...5")|>
  slice(-1)
test_data_fixed
```

Looks pretty good, I think, but what I did won't translate nicely into the function I need to write next. Just wanted to develop a sense for how to do it.

### Second bullet

Now, we need to work on getting a function that takes what's returned by GET() and turns it into a nice tibble Note that I also found the "set_names" function online, which means I can fix the column names all at once (instead of piecemeal).

```{r}
#Functionalizing GET()
make_tib<-function(x){
  query_parsed<-fromJSON(rawToChar(x$content))
  query_data<-as_tibble(query_parsed,.name_repair="unique_quiet")|>
    set_names(query_parsed[1,])|>
    slice(-1)
query_data  
}

```

Note that test is still a defined GET() output from before. Let's take it for a spin.

```{r}
#Trying function
make_tib(test)
```

Seems alright!

### Third bullet
#### Query Validation

Now, the big one. We need to write a function that takes, as inputs, year of survey (default of 2024), numeric variables (default to AGEP, with PWGTP always returned), categorical variables (at least one, with SEX as the default), geography level (default to STATE) and one specific value of the variable.

I'm going to start by  setting up the function defaults and creating error messages if/when improper calls are made. For now, I'm going to just make the function add 1 to a number so I can focus on the function setup (with a focus on functionality to come).

```{r}
#Setting up query validation function. Note "x" and "x+1" there to give funciton something to do besides just check inputs.
query_validation<-function(yr=2024,num_vars="AGEP",cat_vars="SEX",geo_level="STATE",geo_vars="ALL",x=5){
  yr_called<-var_lists[[paste0("vars",yr)]]
  if(yr!=2024 & yr!=2023 & yr != 2022 & yr != 2021){
    stop("year error: You must enter year value of 2021, 2022, 2023 or use default of 2024.")
  }
  if(sum(num_vars %in% c("AGEP", "GASP", "GRPIP", "JWAP","JWDP","JWMNP"))!=length(num_vars) ){
    stop("num_vars error: You must enter a subset of the following numerical variables: AGEP, GASP, GRPIP, JWAP, JWDP or JWMNP.")
  }
  if(sum(cat_vars %in% c("FER", "HHL", "SCH", "SCHL","SEX"))!=length(cat_vars) || length(cat_vars)==0){
    stop("car_vars error: You must enter a nonempty subset of the following categorical variables: FER, HHL, SCH, SCHL,SEX.")
  }
  if(geo_level=="ST"){geo_level<-"STATE"}
  if(length(geo_level)>1 || geo_level %in% c("STATE","REGION","DIVISION") != 1){
    stop("geo_level error: You must select one of the following geographical levels: STATE, REGION or DIVISION.")
  }
  if(geo_level == "REGION" && toupper(geo_vars[1]) != "ALL" && !all(as.character(geo_vars) %in% names(yr_called$REGION$values$item))){
    print(unlist(yr_called$REGION$values$item))
    stop("geo_var error: Since geo_level selection is region, you must select one or more of the regions listed above. Be sure to use the provided (numeric) label.For example, set geo_vars=c('1','2') for Northeast and Midwest, where the single quotation marks are replaced with standard double quotation marks.")
  }
  if((geo_level == "STATE" || geo_level == "ST") && toupper(geo_vars[1]) != "ALL" && !all(as.character(geo_vars) %in% names(yr_called$STATE$values$item))){
    print(unlist(yr_called$STATE$values$item))
    stop("geo_var error: Since geo_level selection is state, you must select one or more of the states from the list above. Be sure to use the provided (numeric) label.For example, set geo_vars=c('01','02') for Alabama and Alaska, where the single quotation marks are replaced with standard double quotation marks.")
  }
  if(geo_level == "DIVISION" && toupper(geo_vars[1]) != "ALL" && !all(as.character(geo_vars) %in% names(yr_called$DIVISION$values$item))){
    print(unlist(yr_called$DIVISION$values$item))
    stop("geo_var error: Since geo_level selection is division, you must select one or more of the divisions from the list above. Be sure to use the provided (numeric) label. For example, set geo_vars=c('0','1') for Puerto Rico and New England (Northeast region), where the single quotation marks are replaced with standard double quotation marks.")
  }
  x+1
  return(geo_level)
}
```
#### Helper functions

Alright, it looks like all my validation stuff is working (at least, the logical tests are), so I'll hopefully be able to call it in my main query function. I started by making my function slowly, by first putting it together with a carefully curated URL, then adding in the validation and tibble-making functions. Then, I worked out how to assign the proper values within the categorical variable columns. So, what's presented here is out-of-sync with development. I didn't want to keep showing you the "make_query" function, then a new helper function, then "make_query" over again with the new helper added in. 

It's worth noting here that I had a really hard time getting all this to work out. For example, in constructing the URL, I initially ran into trouble because, when I wanted multiple  numerical and/or categorical variables, it would create a new URL for each value in the variable. It also took me quite a while to get the ability to pick geography level, and then either all or a subset of the variable values within the level selected. I amended the error code above to provided the lists the user could use to find the labels they need. 

In the assigning-levels to the categorical variables, my plan was to start with the easiest columns (any non-time numeric variables), then handle the categorical ones and then end with using an LLM to sort out the time variables. I should note that, for the categorical variables, I got it working with a single year and variable first and then ran into trouble figuring out how to expand that to multiple variables within a loop. The technicality of how things get treated (e.g., what needs quotation marks, what types of things are allowed to follow a $, etc.) was really difficult for me but a helpful email got me going on that. Finally, I did end up relying on AI to take care of the time variables but the nice thing about LLM's is that they're pretty patient with users asking lots of questions about how things work (and explaining things one line at a time).

To re-iterate my process a bit, I got my function working with defaults first, then folded in the validation (which should stop the whole shebang if it fails), then constructed the link, tibble-fied what came back using the functions from before, and then worked on the factors and levels. What you see below misrepresents the order of all that, but I didn't want to keep re-inserting the big function (make_query) every time I added to it. 

```{r}

# Helper to convert a single "HH:MM a.m. to HH:MM a.m." string to its midpoint time
parse_time_midpoint <- function(time_str) {
  # Return NA for missing, NA strings, or non-range values (e.g., "N/A", "Did not work")
  if (is.na(time_str) || !grepl(" to ", time_str, ignore.case = TRUE)) {
    return(as_hms(NA))
  }
  
  # Clean up Census 'a.m.'/'p.m.' formatting for lubridate parse_date_time
  cleaned_str <- tolower(gsub("\\.", "", time_str))
  parts <- unlist(strsplit(cleaned_str, " to "))
  
  # Parse start and end as date-times on a dummy date. NOte that _dt is really talking about a time here, not a date.
  start_dt <- parse_date_time(parts[1], orders = "I:M p")
  end_dt   <- parse_date_time(parts[2], orders = "I:M p")
  
  # Handle overnight wraparound (e.g., 11:50 pm to 12:05 am)
  if (!is.na(start_dt) && !is.na(end_dt) && end_dt < start_dt) {
    end_dt <- end_dt + days(1)
  }
  
  # Calculate midpoint time. Does calculations on times. Nice.
  midpoint_dt <- start_dt + (end_dt - start_dt) / 2
  
  # Extract time component as hms object
  as_hms(midpoint_dt)
}

# Vectorized version for use inside mutate(). This part showed me a potential work around for something I ran into before: R not running logical tests across vectors. Also, AI stepped me through the process of parsing the values once, then applying them to the values that show up.

time_midpoint <- function(vec) {
  #Identifies unique strings in the input vector
  unique_vals <- unique(vec[!is.na(vec)])
  # Applies the parsing function across the vector (once)
  parsed_unique <- vapply(unique_vals, parse_time_midpoint, FUN.VALUE = hms(0))
  # Creates a vector which is used to match up times and mid-points
  names(parsed_unique) <- unique_vals
  unname(parsed_unique[as.character(vec)])
}
```


```{r}
fix_cols<-function(x,yr,cat_vars=NULL){
  colnames(x)<-toupper(colnames(x))
  # Setting ST to STATE, if necessary
  if ("ST" %in% colnames(x) && !"STATE" %in% colnames(x)) {
    colnames(x)[colnames(x) == "ST"] <- "STATE"
  }
  # Making numeric values numeric.
    num_list <- c("PWGTP", intersect(colnames(x), c("AGEP", "GASP", "GRPIP","JWMNP")))
  for (col in num_list) {
    x[[col]] <- as.numeric(x[[col]])
  }
  
  # Setting levels for categorical (and geographical) variables
 cat_list<-intersect(colnames(x),c("FER","HHL","SCH","SCHL","SEX","STATE", "REGION", "DIVISION"))
 
 yr_called<-var_lists[[paste0("vars",yr)]]
 
 #Assigning levels to categorical (and geographical) variables
 for(var_called in cat_list){
   # Get the right levels for right variable list
   if (!is.null(yr_called[[var_called]]$values$item)) {
   lvls<-unlist(yr_called[[var_called]]$values$item)
   new_name<-paste0(var_called,"_F")
   x[[new_name]]<-factor(lvls[as.character(x[[var_called]])],levels=unname(lvls))
   }
 }
   
 
 # Fix time variables. 
 time_list=intersect(colnames(x),c("JWAP","JWDP"))
 for(var_called in time_list){
   # Get the right levels for right variable list
   if (!is.null(yr_called[[var_called]]$values$item)) {
   lvls<-unlist(yr_called[[var_called]]$values$item)
   # Gets values from tibble, feeds them into time_midpoint function
   time_labels <- lvls[as.character(x[[var_called]])]
   x[[var_called]] <- time_midpoint(time_labels)
   }
 }
 
 
 # Move my new columns to the front, after a new YEAR column since we'll need it later
 x[["YEAR"]] <- yr
 select(x, PWGTP, YEAR, ends_with("_F"), any_of(c("JWDP", "JWAP")),everything())
}


```

#### Query Function

```{r}
#Query function, with validation and tibble-making from above.
make_query<-function(yr=2024,num_vars="AGEP",cat_vars="SEX",geo_level="STATE",geo_vars="ALL"){
  query_validation(yr,num_vars,cat_vars,geo_level,geo_vars,x=5)
  geo_level_lower <- tolower(geo_level)
  geo_level_lower<-if(geo_level_lower == "st"){"state"}else{geo_level_lower}
  
  if(length(geo_vars)==1 && toupper(geo_vars) %in% "ALL"){
    geo<-paste0("&for=",geo_level_lower,":*")
  }else{
    geo<-paste0("&for=",geo_level_lower,":",paste(geo_vars,collapse=","))
  }
  link<-paste0("https://api.census.gov/data/",yr,"/acs/acs1/pums?get=",paste(num_vars,collapse=","),",",paste(cat_vars,collapse=","),",PWGTP",geo,"&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
   first_call<-GET(link)
   made_tib<-make_tib(first_call)
   query_tibble<-fix_cols(made_tib,y=yr)
   class(query_tibble)<-c("census",class(query_tibble))
   query_tibble
  #print(link)
}
```
Now, to test it a bit:

```{r}
try<-make_query(geo_level="DIVISION",geo_vars="2",cat_vars=c("FER","SEX"),num_vars = c("GRPIP","GASP","JWAP"),yr=2021)
try
```
