---
title: "Parsing API Data"
format: html
editor: visual
---

First up, I'll try to query the API using the normal method. 

```{r}
test<-GET("https://api.census.gov/data/2021/acs/acs1/pums?get=AGEP,SEX,ST&for=state:*&SCHL=24&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
test
```

```{r}
test_parsed<-fromJSON(rawToChar(test$content))
test_data<-as_tibble(test_parsed,.name_repair = "unique")
```

```{r}
head(test_data)
```


So, it looks like what we got back has the names for the variables in the first row, so we should fix that. 

```{r}
test_data_fixed<-test_data|>
  rename("AGEP"="...1","SEX"="...2","ST"="...3","SCHL"="...4","STATE"="...5")|>
  slice(-1)
test_data_fixed
```

Looks pretty good, I think, but what I did won't translate nicely into the function I need to write next. 

### Second bullet

Now, we need to work on getting a function that takes what's returned by GET() and turns it into a nice tibble Note that I also found the "set_names" function online, which means I can fix the column names all at once (instead of piecemeal).

```{r}
make_tib<-function(x){
  query_parsed<-fromJSON(rawToChar(x$content))
  query_data<-as_tibble(query_parsed)|>
    set_names(query_parsed[1,])|>
    slice(-1)
query_data  
}

```

Note that test is still a defined GET() output from before. Let's give it a shot.

```{r}
make_tib(test)
```

Seems alright!

Now, the big one. We need to write a function that takes, as inputs, year of survey (default of 2024), numeric variables (default to AGEP, with PWGTP always returned), categorical variables (at least one, with SEX as the default), geography level (default to STATE) and one specific value of the variable.

I'm going to start by  setting up the function defaults and creating error messages if/when improper calls are made. For now, I'm going to just make the function add 1 to a number so I can focus on the function setup (with a focus on functionality to come).

```{r}
query_validation<-function(year=2024,num_vars="AGEP",cat_vars="SEX",geo_level="STATE",geo_vars="All",x=5){
  if(year!=2024 & year!=2023 & year != 2022 & year != 2021){
    stop("year error: You must enter year value of 2021, 2022, 2023 or use default of 2024.")
  }
  if(sum(num_vars %in% c("AGEP", "GASP", "GRPIP", "JWAP","JWDP"))!=length(num_vars) ){
    stop("num_vars error: You must enter a subset of the following numerical variables: AGEP, GASP, GRPIP, JWAP, JWDP.")
  }
  if(sum(cat_vars %in% c("FER", "HHL", "SCH", "SCHL","SEX"))!=length(cat_vars) ){
    stop("car_vars error: You must enter a subset of the following categorical variables: FER, HHL, SCH, SCHL,SEX.")
  }
  if(length(geo_level)>1 || geo_level %in% c("STATE","REGION","DIVISION") != 1){
    stop("geo_level error: You must select one of the following geographical levels: STATE, REGION or DIVISION.")
  }
  if(geo_level == "REGION" & !"All" %in% geo_vars & !all(geo_vars %in% unlist(vars24$REGION$values$item))){
    print(unlist(vars24$REGION$values$item))
    stop("geo_var error: You must select one or more of the regions listed above. Be sure to use the provided (numeric) label.")
  }
  if(geo_level == "STATE" & !"All" %in% geo_vars & !all(geo_vars %in% unlist(vars24$STATE$values$item))){
    print(unlist(vars24$STATE$values$item))
    stop("geo_var error: You must select one or more of the states from the list above. Be sure to use the provided (numeric) label.")
  }
  if(geo_level == "DIVISION" & !"All" %in% geo_vars & !all(geo_vars %in% unlist(vars24$DIVISION$values$item))){
    print(unlist(vars24$DIVISION$values$item))
    stop("geo_var error: You must select one or more of the divisions from the list above. Be sure to use the provided (numeric) label.")
  }
  x+1
}
```

