---
title: "Parsing API Data"
format: html
editor: visual
---

First up, I'll try to query the API using the normal method. 

```{r}
test<-GET("https://api.census.gov/data/2021/acs/acs1/pums?get=AGEP,SEX,ST&for=state:*&SCHL=24&key=640ce0de32b08c9ded3cc11874a33ea9200c2c78")
test
```

```{r}
test_parsed<-fromJSON(rawToChar(test$content))
test_data<-as_tibble(test_parsed)
```

```{r}
head(test_data)
```


So, it looks like what we got back has the names for the variables in the first row, so we should fix that. 

```{r}
test_data_fixed<-test_data|>
  rename("AGEP"="V1","SEX"="V2","ST"="V3","SCHL"="V4","STATE"="V5")|>
  slice(-1)
test_data_fixed
```

Looks pretty good, I think, but what I did won't translate nicely into the function I need to write next. 

### Second bullet

Now, we need to work on getting a function that takes what's returned by GET() abd turns it into a nice tibble Note that I also found the "set_names" function online, which means I can fix the column names all at once (instead of piecemeal).

```{r}
make_tib<-function(x){
  query_parsed<-fromJSON(rawToChar(x$content))
  query_data<-as_tibble(query_parsed)|>
    set_names(query_parsed[1,])|>
    slice(-1)
query_data  
}

```

Note that test is still a defined GET() output from before. Let's give it a shot.

```{r}
make_tib(test)
```

Seems alright!

